Accurate Binding of Sodium and Calcium to a POPC Bilayer by Effective Inclusion of Electronic Polarization

### 2018-03-15 Manuscript/SI.tex
Updates for reviwers' comments

\newpage
\section{Electronic continuum correction applied to CHARMM36 model}
In order to demonstrate that ECC can be applied also to other than Lipid14 model,
we scaled the charges and Lennard-Jones radii of headgroup, glycerol backbone and
carbonyl atoms of the CHARMM36 model of POPC~\cite{klauda10}. We used the
scaling factors optimized for Lipid14 in the main text, i.e., 0.8 for charges and 0.89 for Lennard-Jones
radius. Changes of the headgroup order parameters as a function of added ions
and number density profiles along membrane normal from the ECC-CHARMM model are shown 
in Figs. \ref{scaledCHARMM} and \ref{scaledCHARMMdensity}, respectively,
together with the results from CHARMM36 simulations \cite{catte16,charmmPOPC450mMCaClfiles}
and experiments \cite{akutsu81,altenbach84}. As shown previously \cite{catte16},
the binding of Ca$^{2+}$ is significantly overestimated in CHARMM36 simulations.
In ECC-CHARMM simulations, the Ca$^{2+}$ binding affinity 
is significantly reduced (Fig. \ref{scaledCHARMMdensity}) and the headgroup order parameter response to the added CaCl$_2$
 is in better agreement with experiments (Fig. \ref{scaledCHARMM}).
In line with the results in main text,
Fig. \ref{scaledCHARMMcontacts} also shows faster equilibration time
for Ca$^{2+}$ binding in ECC-CHARMM model than in the original CHARMM36.

In conclusion, applying ECC to CHARMM36 simulation
improved the ion binding affinity to a lipid bilayer, even thought the
scaling factors optimized for Lipid14 were used. Also the area per molecule in
ECC-CHARMM simulation without ions (61 $\AA^2$) is only slightly
lower than the experimental value (64.3 $\AA^2$). 
The optimal scaling factors for CHARMM36 are expected to be different
than for Lipid14 because the initial partial charges are not the same
in different models. Our results indicate that the performance of ion
binding in CHARMM36, and probably in other lipid models as well, could
be further improved by finding the optimal scaling factors for the partial
charges and Lennard-Jones radii.

\newpage

\begin{figure}[!h]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OP.pdf}
  \caption{\label{scaledCHARMM}
    Changes of the headgroup order parameters as a function of NaCl (left) and CaCl$_2$ (right)
    from simulations with CHARMM36 \cite{klauda10}, simulations with ECC-CHARMM parameters and experiments \cite{akutsu81,altenbach84}. 
    CHARMM36 force field parameters for simulations with and without NaCl were taken from CHARMM-GUI \cite{lee15}.
    These were ran 320 ns and 280 ns, respectively, and the first 100 ns was omitted from the analysis.
    Results for CHARMM36 with CaCl$_2$ were calculated from the last 100 ns from the data available at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    ECC-CHARMM simulations with NaCl and CaCl$_2$ were ran 260 ns and 600 ns, respectively, and the first 100 ns were omitted from the analysis.
    Sodium, calcium, and chloride ions were modeled as ECC-ions in these simulations with parameters from Refs. \citenum{martinek17, Pluharova2014}.
    All the other simulation parameters in simulations in this figure were equal to the available dataset at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    The temperature in all simulations was set to 310 K.
  }
\end{figure}

>>>>

\newpage
\section{Electronic continuum correction applied to CHARMM36 model}
In order to demonstrate that the ECC method can be applied also to other models than Lipid14,
we scaled the charges and Lennard-Jones radii of headgroup, glycerol backbone and
carbonyl atoms of the CHARMM36 model of POPC~\cite{klauda10}. We used the
scaling factors optimized for Lipid14 in the main text, i.e., 0.8 for charges and 0.89 for Lennard-Jones
radii. Changes of the headgroup order parameters as functions of added ions
and number density profiles along membrane normal from the ECC-CHARMM model are shown 
in Figs. \ref{scaledCHARMM} and \ref{scaledCHARMMdensity}, respectively,
together with the results from CHARMM36 simulations \cite{catte16,charmmPOPC450mMCaClfiles}
and experiments \cite{akutsu81,altenbach84}. As shown previously \cite{catte16},
the binding of Ca$^{2+}$ is significantly overestimated in CHARMM36 simulations.
In ECC-CHARMM simulations, the Ca$^{2+}$ binding affinity 
is significantly reduced (Fig. \ref{scaledCHARMMdensity}) and the headgroup order parameter response to the added CaCl$_2$
is in a better agreement with experiments than for the original CHARMM model (Fig. \ref{scaledCHARMM}).
In line with the results in main text,
Fig. \ref{scaledCHARMMcontacts} also shows faster equilibration times
for Ca$^{2+}$ binding in the ECC-CHARMM model than in the original CHARMM36.

In conclusion, applying ECC to CHARMM36 simulation
improved the ion binding affinity to a lipid bilayer, even thought the
scaling factors optimized for Lipid14 were used. Also the area per molecule in
the ECC-CHARMM simulation without ions (61 $\AA^2$) is only slightly
lower than the experimental value (64.3 $\AA^2$) \cite{kucerka11}. 
The optimal scaling factors for CHARMM36 are expected to be somewhat different
than for Lipid14 because the initial partial charges are not the same
in different models. Our results indicate that the performance of ion
binding in ECC-CHARMM36, and probably in other lipid models as well, could
be further improved by finding the optimal scaling factors for the partial
charges and Lennard-Jones radii.

It should be noted that the calcium binding to POPC bilayer within the ECC-CHARMM model
is compared here to the CHARMM36 simulation, which was originally done in a 
previous study \cite{catte16,charmmPOPC450mMCaClfiles}.
After the latter simulation has been performed, specific interaction parameters
between calcium and oxygen \cite{kim16} have been introduced in CHARMM-gui
to reduce the binding of calcium to lipid bilayers. These parameters
reduce the calcium binding to a POPC bilayer to the same level as that of sodium (data not shown).
Further analyses of ion binding with different parameters (including the newly modified CHARMM) are in progress
within the NMRlipids IV project (\url{nmrlipids.blogspot.fi}), being beyond the scope of this work.

%\newpage

\begin{figure}[!h]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OP.pdf}
  \caption{\label{scaledCHARMM}
    Changes of the headgroup order parameters as a function of NaCl (left) and CaCl$_2$ (right)
    from simulations with CHARMM36 \cite{klauda10}, simulations with ECC-CHARMM parameters and experiments \cite{akutsu81,altenbach84}. 
    CHARMM36 force field parameters for simulations with and without NaCl were taken from CHARMM-gui \cite{lee15}.
    These were ran 320 ns and 280 ns, respectively, and the first 100 ns was omitted from the analysis.
    Results for CHARMM36 with CaCl$_2$ were calculated from the last 100 ns from the data available at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    ECC-CHARMM simulations with NaCl and CaCl$_2$ were ran 260 ns and 600 ns, respectively, and the first 100 ns were omitted from the analysis.
    Sodium, calcium, and chloride ions were modeled as ECC-ions in these simulations with parameters from Refs. \citenum{martinek17, Pluharova2014}.
    All the other simulation parameters in simulations in this figure were equal to the available dataset at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    The temperature in all simulations was set to 310 K.
    The simulation data is available at Refs. \citenum{charmm36popcT310K,charmm36popc940mMNaClT310K,ECCcharmm36popcT310K,ECCcharmm36popc940mMNaClT310K,ECCcharmm36popc450mMCaClT310K}.
  }
\end{figure}

### 2018-03-15 Manuscript/manuscript.tex
Updates based on reviewers comments
Updates for reviwers' comments

Here we develop a ECC-POPC lipid model that accurately describes binding 
of sodium and calcium ions to the POPC  lipid bilayer. 
The Lipid14~\cite{dickson14} force field 
(available in a Gromacs format from Ref. \citenum{lipid14files}) was used as a starting 
point since 
it provides the most accurate response of the head groups to ions among the available 
lipid models (see Figs.~2~and~5 in Ref.~\citenum{catte16}). Additionally, the Lipid14 model 
provides relatively realistic head group, glycerol backbone, and acyl chain structures~\cite{dickson14,botan15}.
We applied the ECC correction 
to the Lipid14 model of POPC by scaling 
partial charges of the head group, glycerol 
backbone, and carbonyl regions. 
These are the polar parts of phospholipids which can 
contribute to the cation binding.
The similar approach is applied  also to the CHARMM36 model of POPC~\cite{klauda10} in SI.

@@

The order parameters of the $\alpha$ and $\beta$ carbons in the head group are slightly larger in the ECC-POPC model than in the Lipid14 model, which is apparently related to the P-N vector orienting by about 7$^{\circ}$ more toward the water phase in the former model, see Fig.~\ref{OrderParameterCHANGESsurf}. While both models perform relatively well, considering the available experimental evidence, it is not possible to decide which of the two models provides more realistic head group orientations. The ECC-POPC model gives the $\beta$ carbon order parameter value closer to experiments than the Lipid14 model, while the opposite is true for the $\alpha$ carbon. The accuracy of both models in the glycerol backbone region is comparable to other state-of-the-art lipid model available in literature \cite{botan15}, see Fig.~\ref{simVSexpNOions}.
The electrostatic potential of lipid bilayer is compared between ECC-POPC and Lipid14 simulations in SI.

@@

In this study, we employed the electrometer concept to demonstrate that the binding of \ce{Na^+} and \ce{Ca^2+} ions to a POPC lipid bilayer can be accurately described within a force field MD simulation, provided that electronic polarization is implicitly included via the electronic continuum correction~\cite{leontyev11}. This ECC-POPC model is built upon the Lipid14 POPC model~\cite{dickson14} by scaling the partial charges by a factor of 0.8 and reducing the Lennard-Jones radii by a factor of 0.89 for the headgroup, glycerol backbone, and carbonyl atoms. While the structural details of a POPC lipid bilayer in pure water simulated with the newly developed ECC-POPC model agree with experiments with an accuracy comparable to the other state of the art lipid models, the new model in addition reproduces the experimental lipid head group order parameter responses to a cationic surfactant, NaCl, and CaCl$_2$ at varying concentrations. It thus represents a significant improvement over currently available lipid models, which tend to overestimate cation binding affinities~\cite{catte16}.
Similar improvement is demonstrated for the POPC model of CHARMM36 \cite{klauda10} in SI.

>>>>

Here we develop a ECC-POPC lipid model that accurately describes binding 
of sodium and calcium ions to the POPC  lipid bilayer. 
The Lipid14~\cite{dickson14} force field 
(available in a Gromacs format from Ref. \citenum{lipid14files}) was used as a starting 
point since 
it provides the most accurate response of the head groups to ions among the available 
lipid models (see Figs.~2~and~5 in Ref.~\citenum{catte16}). Additionally, the Lipid14 model 
provides relatively realistic head group, glycerol backbone, and acyl chain structures~\cite{dickson14,botan15}.
We applied the ECC correction 
to the Lipid14 model of POPC by scaling 
partial charges of the head group, glycerol 
backbone, and carbonyl regions. 
These are the polar parts of phospholipids which can 
contribute to the cation binding.
A similar approach is also applied to the CHARMM36 model of POPC~\cite{klauda10} in SI.

@@

The order parameters of the $\alpha$ and $\beta$ carbons in the head group are slightly larger in the ECC-POPC model than in the Lipid14 model, which is apparently related to the P-N vector orienting by about 7$^{\circ}$ more toward the water phase in the former model, see Fig.~\ref{OrderParameterCHANGESsurf}. While both models perform relatively well, considering the available experimental evidence, it is not possible to decide which of the two models provides more realistic head group orientations. The ECC-POPC model gives the $\beta$ carbon order parameter value closer to experiments than the Lipid14 model, while the opposite is true for the $\alpha$ carbon. The accuracy of both models in the glycerol backbone region is comparable to other state-of-the-art lipid model available in literature \cite{botan15}, see Fig.~\ref{simVSexpNOions}.
The electrostatic potentials as obtained in ECC-POPC and Lipid14 simulations are compared in the SI.

@@

In this study, we employed the electrometer concept to demonstrate that the binding of \ce{Na^+} and \ce{Ca^2+} ions to a POPC lipid bilayer can be accurately described within a force field MD simulation, provided that electronic polarization is implicitly included via the electronic continuum correction~\cite{leontyev11}. This ECC-POPC model is built upon the Lipid14 POPC model~\cite{dickson14} by scaling the partial charges by a factor of 0.8 and reducing the Lennard-Jones radii by a factor of 0.89 for the headgroup, glycerol backbone, and carbonyl atoms. While the structural details of a POPC lipid bilayer in pure water simulated with the newly developed ECC-POPC model agree with experiments with an accuracy comparable to the other state of the art lipid models, the new model in addition reproduces the experimental lipid head group order parameter responses to a cationic surfactant, NaCl, and CaCl$_2$ at varying concentrations. It thus represents a significant improvement over currently available lipid models, which tend to overestimate cation binding affinities~\cite{catte16}.
A similar improvement using the ECC procedure is demonstrated for the POPC model of CHARMM36 \cite{klauda10} in SI.
